**Bayesian statistics for longitudinal studies in biomedical research**

### 2020-11-17 Manuscripts/Manuscript_AM_v1.Rmd
Updated

When the aforementioned issues arise, rm-ANOVA requires exclusion of all subjects with missing observations from the analysis, thereby increasing costs for the study if the desired statistical power is not met with the remaining observations as it makes necessary to enroll more subjects; and  raising the possibility that the rejection of those partial observations limits the demonstration of significant differences between groups. Additionally, rm-ANOVA uses a _post hoc_ analysis to assess the relevance between the measured response in different groups. This analysis is based on multiple repeated comparisons to estimate a _p-value_, a metric that is widely used as a measure of significance. Because the p-value is highly variable [@halsey2015;@nuzzo2014], multiple comparisons can inflate the false positivity rate [@liu2010],consequently biasing the conclusions of the study.

During the last decade, the biomedical community has started to recognize the limitations of a rm-ANOVA approach in the analysis of longitudinal information. This is exemplified by the pioneering use of  linear mixed effects models (LMEMs)  in certain groups to analyze tumor longitudinal data [@skala2010;@vishwanath2009]. Briefly, these models incorporate _fixed effects_, which correspond to the levels of experimental factors in the study (e.g. the different drug regimens in a clinical trial), and _random effects_, which account for random variation within the population [@pinheiro2006].These models are more flexible than rm-ANOVA as they can accommodate missing observations for multiple subjects, and allow different modeling strategies for the variability within each measure in every subject [@west2014;@pinheiro2006].On the other hand, they impose restrictions in the distribution of the errors of the model and random effects [@gueorguieva2004;@schielzeth2020].

Additionally, another assumption made in both rm-ANOVA and LMEMs models is that a linear relationship is expected between the observed response and the covariates across the study [@pinheiro2006].This common assumption to both rm-ANOVA and LMEMs causes the models to be restrictive in their inferences when used in longitudinal data that does not follow a linear trend. In biomedical research, this particular behavior in longitudinal has been reported in studies of tumor response to radio/chemotherapy in preclinical and clinical settings[@vishwanath2009; @roblyer2011;@tank2020;@skala2010;@demidov2018], and wound healing and metabolism[@jones2018;@grice2010;@young1994]. These studies have shown that that the collected signal does not follow a linear trend over time, and presents high variability at different time points, making the estimations of a LMEM or rm-ANOVA model inconsistent with the pattern of the observed variations. Additionally,  although it is possible that a post hoc analysis is able to find “significance”( _p-value_<0.05) by using multiple comparisons between the model terms, this estimator would be inherently biased because of the lack of fit between the information and the model.

@@

\begin{equation}
(\#eq:labelANOVA)
y_{kit}=\mu+\gamma_k+\tau_t+(\gamma \tau)_{kt}+\pi_{i(k)}+e_{kit}
\end{equation}

@@

In this model, $\mu$ represents the group mean, $\gamma_h$ is the _fixed effect_ of group $h$, $\tau_j$ is the fixed effect of time $j$, and $(\gamma \tau)_{hj}$ represents the interaction of time and group effects. The term $\pi_{ij}$ represents the _random effects_ for each subject within each group.Finally, $e_{hij}$ represents the independent random error terms, which need to be normally distributed with mean 0 [@davis2002]. The model then, is a linear combination of terms, and if plotted, it would a straight line.

@@

In a longitudinal study, there is a relationship between repeated measures on the same subject. In other words, there is an expected _variance_ between those measurements, and since repeated measures are also taken on the rest of the subjects within each group, there is a _covariance_ between  measurements at each time point within each group. The _covariance matrix_ (also known as the variance-covariance matrix) is a matrix that captures the variation between and within subjects in a longitudinal study[@wolfinger1996] (For an in-depth analysis of the covariance matrix see [@west2014;@weiss2005]).

For practical purposes, an rm-ANOVA analysis assumes that the covariance matrix has a specific construction known as _compound symmetry_ (also known as "sphericity" or "circularity"). This construction implies that the between-subject variance and within-subject correlation  are constant across time [@weiss2005;@geisser1958;@huynh1976]. However, it has been shown that this condition is is frequently unjustified [@maxwell2017] because the correlation between measurements tends to change over time; and it is higher between consecutive measurements [@gueorguieva2004;@ugrinowitsch2004].When the data violates the sphericity assumption, the false positivity rate is inflated [@haverkamp2017].

In the case of LMEMs, they have the advantage of allowing different structures of the variance-covariance matrix [@pinheiro2006].

@@

In a longitudinal study, missing observations are an issue that arises frequently. In biomedical research, this situation can be caused by reasons beyond the control of the investigator [molenberghs2004].Dropout from patients, or attrition or injury in animals are among the reasons for missing observations. Statistically, missing information can be classified as _missing at random_ (MAR), _missing completely at random_ (MCAR), and _missing not at random_ (MNAR) [@weiss2005].  In a MAR scenario, the pattern of the missing information is related to some variable in the dataset, but it is not related to the variable of interest [@scheffer2002]. If the data are MCAR, this means that the missigness is completely unrelated to the collected information[@potthoff2006], and in the case of MNAR the missing values are dependent on their value. In the case of LMEMs, one key advantage over rm-ANOVA is that the model can work with missing observations that are MAR, whereas rm-ANOVA needs to

*talk about missing at random, mcar etc*

#### 1.2 the case of GMEMs using splines and how they work and how they are better than rm-ANOVA

>>>>

When the aforementioned issues arise, rm-ANOVA requires exclusion of all subjects with missing observations from the analysis, thereby increasing costs for the study if the desired statistical power is not met with the remaining observations as it makes necessary to enroll more subjects [@ma2012]; and  raising the possibility that the rejection of those partial observations limits the demonstration of significant differences between groups. Additionally, rm-ANOVA uses a _post hoc_ analysis to assess the relevance between the measured response in different groups. This analysis is based on multiple repeated comparisons to estimate a _p-value_, a metric that is widely used as a measure of significance. Because the p-value is highly variable [@halsey2015;@nuzzo2014], multiple comparisons can inflate the false positivity rate [@liu2010],consequently biasing the conclusions of the study.

During the last decade, the biomedical community has started to recognize the limitations of a rm-ANOVA approach in the analysis of longitudinal information. This is exemplified by the use of  linear mixed effects models (LMEMs)  in certain groups to analyze tumor longitudinal data [@skala2010;@vishwanath2009]. Briefly, these models incorporate _fixed effects_, which correspond to the levels of experimental factors in the study (e.g. the different drug regimens in a clinical trial), and _random effects_, which account for random variation within the population [@pinheiro2006].These models are more flexible than rm-ANOVA as they can accommodate missing observations for multiple subjects, and allow different modeling strategies for the variability within each measure in every subject [@west2014;@pinheiro2006].On the other hand, they impose restrictions in the distribution of the errors of the model and random effects [@gueorguieva2004;@schielzeth2020].

Additionally, another assumption made in both rm-ANOVA and LMEMs models is that a linear relationship exists between the observed response and the explanatory variables (covariates) across the study [@pinheiro2006].This common assumption to both rm-ANOVA and LMEMs causes the models to be restrictive in their inferences when used in longitudinal data that does not follow a linear trend. In biomedical research, this particular behavior in longitudinal has been reported in studies of tumor response to radio/chemotherapy in preclinical and clinical settings[@vishwanath2009; @roblyer2011;@tank2020;@skala2010;@demidov2018], and wound healing and metabolism[@jones2018;@grice2010;@young1994]. These studies have shown that that the collected signal does not follow a linear trend over time, and presents high variability at different time points, making the estimations of a LMEM or rm-ANOVA model inconsistent with the pattern of the observed variations. Additionally,  although it is possible that a _post hoc_ analysis is able to find “significant” results ( _p_<0.05) by using multiple comparisons between the model terms, this metric is inherently biased due to the discrepancy between the data and the model.

@@

\begin{equation}
\label{eq:ANOVA}
$$y_{kit}=\mu+\gamma_k+\tau_t+(\gamma \tau)_{kt}+\pi_{i(k)}+e_{kit}$$
\end{equation}

@@

In this model, $\mu$ represents the group mean, $\gamma_h$ is the _fixed effect_ of group $h$, $\tau_j$ is the fixed effect of time $j$, and $(\gamma \tau)_{hj}$ represents the interaction of time and group effects. The term $\pi_{ij}$ represents the _random effects_ for each subject within each group.Finally, $e_{hij}$ represents the independent random error terms, which need to be normally distributed with mean 0 [@davis2002]. The model then, is a linear combination of terms, and if plotted, represents a line.

@@

In a longitudinal study there is an expected _variance_ between repeated measurements on the same subject, and since repeated measures are also taken on the rest of the subjects within each group, there is a _covariance_ between  measurements at each time point within each group. The _covariance matrix_ (also known as the variance-covariance matrix) is a matrix that captures the variation between and within subjects in a longitudinal study[@wolfinger1996] (For an in-depth analysis of the covariance matrix see [@west2014;@weiss2005]).

For practical purposes, an rm-ANOVA analysis assumes that the covariance matrix has a specific construction known as _compound symmetry_ (also known as "sphericity" or "circularity"). Under this assumption, the between-subject variance and within-subject correlation  are constant across time [@weiss2005;@geisser1958;@huynh1976]. However, it has been shown that this condition is frequently unjustified because the correlation between measurements tends to change over time [@maxwell2017]; and it is higher between consecutive measurements [@gueorguieva2004;@ugrinowitsch2004].When the data violates the sphericity assumption, the false positivity rate is inflated [@haverkamp2017].

In the case of LMEMs, one of their main advantages is that they allow multiple structures for the variance-covariance matrix including exponential, autoregressive of order 1, rational quadratic and others[@pinheiro2006]. The choice for the covariance matrix needs to be carefully considered by the investigator based on the analyzing the collected data.

@@

Missing observations are an issue that arises frequently in longitudinal studies. In biomedical research, this situation can be caused by reasons beyond the control of the investigator [molenberghs2004].Dropout from patients, or attrition or injury in animals are among the reasons for missing observations. Statistically, missing information can be classified as _missing at random_ (MAR), _missing completely at random_ (MCAR), and _missing not at random_ (MNAR) [@weiss2005].  In a MAR scenario, the pattern of the missing information is related to some variable in the dataset, but it is not related to the variable of interest [@scheffer2002]. If the data are MCAR, this means that the missigness is completely unrelated to the collected information [@potthoff2006], and in the case of MNAR the missing values are dependent on their value. An rm-ANOVA model assumes complete observations for all subjects, and therefore subjects with one or more missing observations are excluded from the analysis. This is inconvenient because the remaining subjects might not accurately represent the population, and statistical power is affected by this reduction in sample size [@ma2012].

In the case of LMEMs, one key advantage over rm-ANOVA is that inferences from the model are valid when missing observations in the dataset exist that are MAR [@west2014]. The pattern of missing observations can be considered MAR if the missing observations are not related any of the other variables measured in the study [@maxwell2017]. For example, if attrition occurs in all mice that had lower weights at the beginning of a chemotherapy response study. Because in this case the missigness is unrelated to other variables of interest, the missing data can be considered MAR.

This section has presented the limitations of rm-ANOVA to analyze longitudinal information and the differences of LMEMs with regards to missing data and the modeling of the covariance matrix. Of notice, LMEMs offer a more robust and flexible approach than rm-ANOVA and if the data follows a linear trend, they provide an excellent choice to derive inferences from a repeated measures study. However, when the data presents high variability, LMEMs fail to capture the non-linear trend of the data. To analyze such type of data, we propose the use of generalized additive models (GAMs) in the following section.

#### 2 GAMs as a special case of Generalized Linear Models:

A GAM is a special case of the Generalized Linear Model (GLM), a framework that allows for response distributions that do not follow a normal distribution [@wood2017;@hastie1987].  [@simpson2018].  To allow more flexibility, GAMs use _smooth functions_  to model the relationship between the covariates and the response. The smooth functions are represented using splines over  evenly spaced ranges of the covariates known as _knots_.  A GAM model can be represented as:

\begin{equation}
(\#eq:labelGAM)
y_{t}=\beta_0+f(x_t)+\epsilon_t  
\end{equation}

Where $y_t$ is the response at time $t$, $\beta_0$ is the expected value at time 0, the change of $y$ over time is represented by the function $f(x_t)$ and $\epsilon_t$ represents the residuals.

*This equation above does not have an "interaction term" similar to that of anova.*

This model is similar in its construction to that of \ref{eq:ANOVA}, but the key difference relies on the fact that instead of a linear relationship for the covariates, a function is used.
